---
title: "Audit & Diagnostics"
subtitle: "Algorithmic Health Checks, Autocorrelation, and Environment Manifest"
---

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(ggplot2)
library(gt)
library(reactable)
library(scales)
library(tidyverse)

source("R/metropolis_sampler.R")
source("R/risk_functions.R")
source("R/plots.R")

# 1. Ingest fixture and run two independent chains to test mixing
model_fixture <- build_risk_target("data/synthetic_returns.csv")
d <- model_fixture$d
asset_names <- model_fixture$asset_names

init_theta1 <- rep(0, d)
names(init_theta1) <- asset_names

init_theta2 <- rep(0.001, d)
names(init_theta2) <- asset_names

chain1 <- run_metropolis(
  target_log_posterior = model_fixture$target_fn,
  initial_theta = init_theta1,
  iterations = 3000,
  proposal_sd = 0.00032,
  seed = 42
)

chain2 <- run_metropolis(
  target_log_posterior = model_fixture$target_fn,
  initial_theta = init_theta2,
  iterations = 3000,
  proposal_sd = 0.00032,
  seed = 99
)

# Burn-in discard (first 500 draws)
draws1_clean <- chain1$draws %>% slice_tail(n = 2500)
draws2_clean <- chain2$draws %>% slice_tail(n = 2500)

# 2. Compute Robust Effective Sample Size (ESS) using Geyer's initial sequence estimator
ess_tech  <- compute_ess_robust(draws1_clean$Tech_Growth)
ess_treas <- compute_ess_robust(draws1_clean$US_Treas_10Y)
ess_large <- compute_ess_robust(draws1_clean$US_LargeCap)

# Diagnostic inline variables
txt_acc_c1   <- sprintf("%.1f%%", chain1$acceptance_rate * 100)
txt_acc_c2   <- sprintf("%.1f%%", chain2$acceptance_rate * 100)
txt_ess_tech <- comma(ess_tech)
txt_n_draws  <- comma(nrow(draws1_clean))

```

### Introduction
In institutional risk management, stochastic simulation models must be transparent and auditable. When using Markov Chain Monte Carlo to evaluate high-dimensional portfolio distributions, point estimates alone are insufficient. Because MCMC generates a sequence of dependent parameter proposals rather than independent samples, model validation requires empirical verification of convergence, parameter space exploration, and statistical sample efficiency.

This audit page examines the algorithmic health checks for the 10-dimensional Gaussian random walk. We evaluate multi-chain convergence, measure autocorrelation decay across factor lags, and apply Geyer's initial sequence estimator to compute robust Effective Sample Size (ESS). This diagnostic audit outlines both the utility of Random-Walk Metropolis and the operational limitations encountered when isotropic proposals are applied to correlated factor distributions.

### Dataset Architecture & Methodological Rationale
The analysis utilizes a fixed synthetic fixture of 250 trading days across 10 correlated asset classes (data/synthetic_returns.csv), generated from a known multivariate normal distribution ‭$\mathcal{N}(\boldsymbol{\mu}, \mathbf{\Sigma})$‬‭‬‭‬‭‬ ‭‬‭‬ calibrated to realistic risk levels: annualized returns between 3.5% and 11.0%, volatilities between 8.0% and 26.0%, and correlations from -0.30 to +0.78.

Using a synthetic baseline provides a controlled ground truth. In empirical market feeds, true data-generating distributions are obscured by regime shifts, structural breaks, and survivorship bias. By defining a synthetic population, we establish an explicit benchmark to test whether the sampler recovers sample parameters. Evaluating 10 factors illustrates the scaling challenge where uniform grid search (‭$10^{10}$‬ points) becomes computationally impractical, while 250 observations preserve realistic parameter standard errors for downside modeling.

```{r retrieve-data}

synthetic_data_returns <- as_tibble(read_csv("data/synthetic_returns.csv", show_col_types = FALSE))

reactable(synthetic_data_returns,
  defaultPageSize = 10,
  highlight   = TRUE,
  striped     = TRUE,
  resizable   = TRUE,
  defaultColDef = colDef(
    align = "left",
    format = colFormat(digits = 7),
    width = 110
  ),
  columns = list(
    trading_day = colDef(align = "center", format = colFormat(digits = 0))
  ),
  class = ".react_table",
  theme = reactableTheme(
    borderColor = "#dfe2e5",
    stripedColor = "#f6f8fa",
    style = list(fontSize = "13px", fontFamily = "Roboto, Helvetica, Arial, sans-serif"),
    searchInputStyle = list(width = "100%")
  )
)

```

### The Diagnostic Framework
In quantitative validation workflows, evaluating an MCMC sampler involves three core criteria:
	1.	Stationarity: The chain has converged to the target distribution and is invariant to initialization coordinates.
	2.	Mixing Across Dimensions: The chain traverses parameter space without prolonged freezing or sluggish diffusion.
	3.	Effective Sample Size (ESS): Autocorrelation decays rapidly enough to provide sufficient independent information for quantile estimation (Vehtari et al., 2021).


### Multi-Chain Mixing Verification
To evaluate convergence, two independent chains are tracked from distinct initialization vectors (‭$\boldsymbol{\theta}^{(0)}_1 = \mathbf{0}$‬‭‬ vs. ‭$\boldsymbol{\theta}^{(0)}_2 = 0.001 \cdot \mathbf{1}$‬‭‬‭‬).

```{r multi-chain-trace, fig.cap = "Figure 1: Two independent MCMC chains tracking Tech Growth expected returns"}

#| fig-width: 10
#| fig-height: 8

trace_df <- bind_rows(
  chain1$draws %>% select(iteration, Tech_Growth) %>% mutate(Chain = "Chain 1 (Seed 42)"),
  chain2$draws %>% select(iteration, Tech_Growth) %>% mutate(Chain = "Chain 2 (Seed 99)")
)

ggplot(trace_df, aes(x = iteration, y = Tech_Growth, color = Chain)) +
  geom_line(alpha = 0.75, linewidth = 0.5) +
  geom_vline(xintercept = 500, linetype = "dashed", color = "#c0392b", linewidth = 0.8) +
  annotate("text", x = 520, y = max(trace_df$Tech_Growth) * 0.9, label = "End of Burn-In (500)", fontface = "bold", color = "#c0392b", hjust = 0, size = 3.5) +
  scale_color_manual(values = c("#1b4f72", "#e67e22")) +
  scale_y_continuous(labels = label_percent(accuracy = 0.01)) +
  labs(
    title = "Multi-Chain Stationarity Diagnostic: Tech Growth Exposure",
    subtitle = "Both chains navigate into the stationary posterior distribution",
    x = "Iteration Step",
    y = "Daily Expected Return",
    color = "Sampler Execution"
  ) +
  theme_mcmc()

```

**Diagnostic Review: Multi-Chain Trace**

1. **Stationary Overlap:** Discarding the first 500 iterations eliminates initial burn-in, after which both chains traverse the same posterior band.
2. **Initialization Distance Note:** The initialization vectors (‭$\mathbf{0}$‬ vs. ‭$0.001 \cdot \mathbf{1}$‬‭‬) differ by approximately 25 basis points annualized. While this demonstrates local convergence into the posterior mode, formal model validation typically tests widely separated, overdispersed starting values (e.g., ‭$\pm 3 \cdot \text{SE}$‬‭‬) across 4 or more chains to confirm mixing across distant regions of parameter space.

---

### Autocorrelation Decay Analysis
Because Metropolis-Hastings operates via local proposals, consecutive draws exhibit positive serial dependence.

```{r acf-plot, fig.cap = "Figure 2: Autocorrelation decay across consecutive MCMC lags for selected portfolio risk factors"}

#| fig-width: 10
#| fig-height: 8

calc_acf_df <- function(vec, factor_name, max_lag = 35) {
  res <- acf(vec, lag.max = max_lag, plot = FALSE)
  tibble(
    lag = as.numeric(res$lag),
    autocorrelation = as.numeric(res$acf),
    factor = factor_name
  )
}

acf_combined <- bind_rows(
  calc_acf_df(draws1_clean$Tech_Growth, "Tech Growth"),
  calc_acf_df(draws1_clean$US_Treas_10Y, "US Treasuries (10Y)"),
  calc_acf_df(draws1_clean$US_LargeCap, "US Large Cap")
)

ggplot(acf_combined, aes(x = lag, y = autocorrelation, fill = factor)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0.10, linetype = "dashed", color = "#7f8c8d") +
  scale_fill_manual(values = c("#2c3e50", "#2980b9", "#27ae60")) +
  labs(
    title = "Autocorrelation Decay by Factor",
    subtitle = "Autocorrelation decay profiles highlight the impact of isotropic proposal scaling on correlated parameters",
    x = "Lag Step",
    y = "Autocorrelation Coefficient",
    fill = "Risk Factor"
  ) +
  theme_mcmc()

```

**Diagnostic Analysis: The Proposal-Geometry Mismatch**

The autocorrelation profiles highlight an operational challenge when using an unscaled, isotropic proposal (‭$\sigma_\epsilon \mathbf{I}$‬‭‬) across correlated dimensions:

1. **Treasuries (10Y):** Autocorrelation decays quickly, dropping below 0.50 by lag 7 and reaching baseline levels near lag 20. With low return volatility (8%), the parameter space is compact.
2. **Equity Beta (Tech Growth):** Autocorrelation exhibits high persistence, remaining around 0.65 at lag 35. Because the scalar proposal step (‭$\sigma_\epsilon = 0.00032$‬‭‬) is applied identically across all dimensions, it does not account for the differing scales (Treasuries ‭$\text{SE} \approx 0.03\%$‬‭‬‭‬ vs. Tech ‭$\text{SE} \approx 0.10\%$‬‭‬‭‬) or the correlated geometry of the target density. As a result, the chain traverses the principal variance axes slowly, leading to persistent serial correlation.


---

### Robust Effective Sample Size (ESS) Audit
To avoid truncation bias in persistent chains, Effective Sample Size is computed using Geyer's initial sequence estimator across all post-burn draws:

```{r ess-table}

tibble(
  Asset_Class = c("Equity Beta (Tech Growth)", "Duration Hedge (10Y Treasuries)", "Core Market (US Large Cap)"),
  Total_Post_Burn_Draws = rep(nrow(draws1_clean), 3),
  Effective_Sample_Size = c(ess_tech, ess_treas, ess_large),
  Sampling_Efficiency = c(ess_tech, ess_treas, ess_large) / nrow(draws1_clean),
  Autocorrelation_Lag10 = c(
    acf(draws1_clean$Tech_Growth, plot = FALSE)$acf[11],
    acf(draws1_clean$US_Treas_10Y, plot = FALSE)$acf[11],
    acf(draws1_clean$US_LargeCap, plot = FALSE)$acf[11]
  )
) %>%
  gt() %>%
  tab_header(
    title = md("**MCMC Chain Efficiency & Effective Sample Size Audit**"),
    subtitle = "Robust Geyer initial-sequence evaluation of chain correlation and sampling capacity"
  ) %>%
  cols_label(
    Asset_Class = "Risk Factor Asset",
    Total_Post_Burn_Draws = "Post-Burn Draws",
    Effective_Sample_Size = "Effective Draws (ESS)",
    Sampling_Efficiency = "Efficiency Ratio",
    Autocorrelation_Lag10 = "Lag-10 Autocorr"
  ) %>%
  fmt_integer(columns = c(Total_Post_Burn_Draws, Effective_Sample_Size)) %>%
  fmt_percent(columns = Sampling_Efficiency, decimals = 1) %>%
  fmt_number(columns = Autocorrelation_Lag10, decimals = 3) %>%
  tab_options(table.font.size = 12)

```


**Model Validation Findings**

A model validation review of the ESS diagnostics yields three key takeaways:

1 **Sampling Efficiency Reflects Correlation Drag:** Out of `{r} txt_n_draws` post-burn iterations, Tech Growth yields `{r} txt_ess_tech` effective independent draws—a sampling efficiency of approximately 1.5% to 2.0%. While the acceptance rate of `{r} txt_acc_c1` falls within standard aggregate ranges, the chain experiences high serial correlation due to isotropic proposal variance. <br/>
2. **Quantile Precision Considerations:** Established Bayesian validation standards (Vehtari et al., 2021) recommend an ESS of at least 300 to 400 draws to produce stable tail quantiles and bound Monte Carlo standard error. While an ESS of ‭$\approx 39\text{--}82$‬‭‬‭‬ draws is sufficient to estimate the posterior mean vector, downstream tail risk quantiles (such as 95% VaR and Expected Shortfall credible intervals) retain non-trivial Monte Carlo variance. <br/>
3. **Paths to Algorithmic Improvement:** This diagnostic demonstrates the trade-offs of first-principles isotropic Random-Walk Metropolis. In production risk environments, proposal efficiency can be improved by:

-   **Proposal Covariance Preconditioning:** Scaling jumps using the empirical covariance matrix ‭$\mathbf{\Sigma}_{\text{prop}} = \frac{2.38^2}{d} \hat{\mathbf{\Sigma}}_{\text{post}}$‬‭‬‭‬ (Roberts & Rosenthal, 2001).
-	 **Adaptive Sampling:** Implementing adaptive Metropolis algorithms to dynamically learn posterior geometry.
-	 **Gradient-Based Methods:** Using Hamiltonian Monte Carlo (HMC) or NUTS to navigate high-dimensional curvature efficiently.

See the exact-posterior benchmark on the Portfolio Risk page for the effect of low ESS on credible-interval width.

---

### Toolchain & Runtime Environment

```
mcmc-in-action/
├── _quarto.yml               # Quarto configuration (Sandstone theme, navbar)
├── index.qmd                 # Intuition & 2D random walk visualization
├── 01-mechanics.qmd          # Metropolis-Hastings engine & tuning diagnostics
├── 02-risk-application.qmd   # 10-factor portfolio risk model & tail estimates
├── 03-reproducibility.qmd    # Convergence diagnostics & computational manifest
├── R/
│   ├── generate_data.R       # Generates synthetic 10-asset returns fixture
│   ├── metropolis_sampler.R  # Pure R Metropolis-Hastings sampler
│   ├── risk_functions.R      # Joint log-posterior & tail risk evaluation
│   └── plots.R               # ggplot2 themes and diagnostic plots
└── data/
    └── synthetic_returns.csv # Frozen 250-day x 10-factor returns matrix
```

#### Session Information

```{r session-info, echo = FALSE}

sessioninfo::session_info()

```

```{css, echo = FALSE}
.react_table {
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  max-height: 600px;
}

```


