---
title: "The Mechanics of Markov Chain Monte Carlo"
subtitle: "Benchmarking First-Principles Parameter Exploration in Correlated Factor Spaces"
author: "Patrick Lefler"
date: "September 24, 2026"
abstract: "Portfolio risk models must estimate many correlated parameters at once, 
and deterministic grid integration becomes impractical as the number of 
dimensions grows (\\(K^D\\)). This project builds a Random-Walk Metropolis 
sampler from first principles in R, with no MCMC libraries, and applies 
it to the expected returns of a 10-factor portfolio spanning equity, 
credit, duration, and commodity exposures. The model is deliberately 
chosen to have an exact answer: with multivariate normal returns and 
covariance fixed at its sample estimate, the posterior is 
\\(\\boldsymbol{\\mu} \\mid \\mathbf{R}, \\mathbf{\\Sigma} \\sim \\mathcal{N}(\\bar{\\mathbf{x}}, \\tfrac{1}{n}\\mathbf{\\Sigma})\\). This lets every MCMC output be checked against closed-form truth.

The results show how a sampler can look healthy and still understate risk. 
At a 28.9% acceptance rate, well within conventional targets, the chain 
reproduces the exact posterior means of Value at Risk (1.05%) and Expected 
Shortfall (1.33%). The diagnostics tell a different story. All ten factors 
are at or above the split-\\(\\hat{R}\\) convergence threshold of 1.01 across 
four chains (maximum 1.057). Effective Sample Size is only 16–65 out of 
2,500 post-burn draws, far below the recommended 400. Most important for 
risk reporting, the MCMC credible interval for VaR is about 17% narrower 
than the exact interval, so the sampler understates parameter uncertainty.

The cause is a single isotropic proposal step applied to factors with very 
different scales and strong correlations. The findings show why acceptance 
rates alone are not sufficient evidence of convergence, and why production 
risk engines need covariance-preconditioned or gradient-based samplers, 
validated with multi-chain diagnostics and, where possible, exact benchmarks."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(dplyr)
library(gt)
library(scales)

source("R/metropolis_sampler.R")
source("R/risk_functions.R")
source("R/plots.R")

# Run toy 2D chain for hero visual
target_bivariate <- function(theta) {
  mu <- c(0.5, 0.5)
  sigma <- matrix(c(1.0, 0.70, 0.70, 1.0), nrow = 2)
  -0.5 * t(theta - mu) %*% solve(sigma) %*% (theta - mu)
}

hero_chain <- run_metropolis(
  target_log_posterior = target_bivariate,
  initial_theta = c(-3.2, 3.0),
  iterations = 300,
  proposal_sd = 0.45,
  seed = 42
)

```

### Introduction
Markov Chain Monte Carlo (MCMC) is a computational class of algorithms designed to sample from complex, high-dimensional probability distributions when analytical integration is intractable. In portfolio risk management, evaluating joint risk across multiple assets requires characterizing dependent parameters simultaneously—including factor means, volatilities, and cross-asset correlations. For non-standard priors or complex likelihoods, deterministic numerical grid integration scales exponentially with dimension (‭$K^D$‬), quickly exceeding practical computational limits.

MCMC bypasses deterministic grid mapping by replacing brute-force evaluation with an informed random walk. Guided by transition rules such as the Metropolis-Hastings criterion, the sampler proposes local moves, accepting higher-density coordinates and probabilistically accepting lower-density proposals. Over extended sampling runs, the chain visits parameter states in proportion to their underlying posterior density, mapping the target space without evaluating intractable normalizing constants.

In this demonstration, we focus on isolating parameter estimation uncertainty. Using a controlled multivariate Gaussian portfolio framework, we deliberately pair our pure-R Metropolis sampler against a model with an exact analytical solution: ‭$\boldsymbol{\mu} \mid \mathbf{R}, \mathbf{\Sigma} \sim \mathcal{N}(\bar{\mathbf{x}}, \frac{1}{n}\mathbf{\Sigma})$‬‭‬‭‬ ‭‬‭‬‭‬‭‬‭‬ ‭‬‭‬‭‬. By holding covariance fixed and evaluating expected returns against this closed-form baseline, we provide an auditable environment to examine convergence, proposal efficiency, and the propagation of parameter uncertainty into Value at Risk (VaR) and Expected Shortfall (ES).

---

### Plain-Language Translation: How It Works
Imagine being placed on a large, pitch-black terrain with only a small flashlight, tasked with mapping the highest peaks and ridges. You cannot view the entire landscape at once, and pacing every square foot on a rigid grid would take an impractical amount of time. Instead, you take incremental random steps. When a step leads uphill toward higher ground, you take it; when a step heads downhill, you occasionally take it anyway to avoid staying confined to a single minor hill.
Because you spend most of your time traversing higher ridges and pass quickly through low valleys, your path naturally traces the contours of the landscape. In financial modeling, algorithms like Markov Chain Monte Carlo deploy this sequential exploration to estimate portfolio risk. Rather than evaluating billions of arbitrary grid points across multiple correlated factors, the sampler concentrates computational effort directly in regions of high probability.

---

### The Dimensional Scaling Challenge
When managing risk across multiple correlated factors—equity betas, credit spreads, duration, and commodities—quantifying joint parameter uncertainty is essential.
In low dimensions, deterministic numerical quadratures or grid evaluations are trivial. However, as the parameter count (‭$D$‬) increases, Cartesian grid evaluations scale exponentially (‭$K^D$‬).

```{r grid-breakdown-table, echo = TRUE}

generate_dimension_comparison() %>%
  gt() %>%
  tab_header(
    title = md("**Dimensional Complexity: Grid Search vs. Stochastic Sampling**"),
    subtitle = "Scaling behavior of uniform grid evaluation vs. stochastic simulation"
  ) %>%
  cols_label(
    Dimensions = "Factor Horizon",
    Grid_Points_Per_Axis = "Points / Axis",
    Total_Grid_Evaluations = "Grid Evaluations ($K^D$)",
    Grid_Feasibility = "Grid Feasibility",
    MCMC_Iterations_Required = "Target MCMC Draws",
    MCMC_Solve_Time = "Observed MCMC Time"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "#c0392b"),
    locations = cells_body(columns = Total_Grid_Evaluations, rows = 2:3)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", color = "#27ae60"),
    locations = cells_body(columns = MCMC_Solve_Time)
  ) %>%
  tab_options(table.font.size = 12)

```

**Low-Dimensional Space (‭$D = 2$‬‭‬):** A coarse 10-point grid requires only ‭$10^2 = 100$‬‭‬ function evaluations, making numerical quadratures instantaneous.

**Moderate Parameter Space (‭$D = 10$‬‭‬):** Uniform evaluation along each axis requires ‭$10^{10}$‬ calculations (10 billion operations)—creating significant runtime bottlenecks in standard computing environments, whereas stochastic sampling traverses the distribution in seconds.

**Full Covariance Horizon (‭$D = 20$‬‭‬):** Modeling 20 factor means alongside full cross-asset covariances involves ‭$D + D(D+1)/2 = 230$‬‭‬‭‬‭‬‭‬‭‬‭‬‭‬ distinct parameters. Evaluating such spaces on uniform deterministic grids is computationally intractable, making stochastic simulation essential for non-conjugate structures.

---

### The Random Walk Engine
Markov Chain Monte Carlo navigates probability distributions via sequential, localized transitions:

1. **Propose a Move:** The chain samples a candidate coordinate ‭$\theta^*$‬ from a proposal distribution centered on the current state ‭$\theta^{(t)}$‬.
2. **Evaluate Density Ratio:** It computes the target posterior probability ratio between candidate and current locations.
3. **Accept or Reject:** Proposals into higher-density regions are accepted; lower-density moves are accepted probabilistically, enabling the chain to explore parameter tails.

::: {.callout-note icon=false}
## The Metropolis-Hastings Random Walk
Evaluating joint portfolio risk across multiple interdependent assets presents substantial computational complexity when models deviate from simple analytical assumptions. Calculating every joint market permutation across a fine grid requires billions of evaluations, creating an operational bottleneck. The Metropolis-Hastings algorithm resolves this by deploying a sequential "digital scout." Rather than attempting to evaluate the entire parameter space uniformly, the algorithm takes incremental steps through potential states. If a proposed move discovers a region of higher probability, the chain transitions there; if it steps toward a lower-probability scenario, it still accepts the move occasionally to ensure thorough exploration.
Over thousands of iterations, the algorithm visits parameter combinations in proportion to their joint posterior probability. This produces an empirical distribution of parameter uncertainty. For executive decision-makers, this turns otherwise intractable high-dimensional estimation into a practical sampling process, providing an empirical baseline for setting risk capital and stress limits.

:::

```{r hero-random-walk-model, fig.cap = "Figure 1: Metropolis-Hastings chain traversing a correlated bivariate parameter space."}

#| fig-width: 10
#| fig-height: 10

plot_random_walk_2d(hero_chain$draws, n_steps = 180)

```

#### Traversal Mechanics
- **Stationary Convergence:** Initializing from an arbitrary coordinate in the outer tail, the chain transitions toward the high-density contour core.
- **Proportional Dwell Time:** Once stationary, the chain continues moving, spending time across regions in proportion to target posterior density.

---

### Project Structure & Roadmap

This project demonstrates the implementation, diagnostic auditing, and tail-risk application of first-principles MCMC:

- [The Metropolis Engine](01-mechanics.qmd): The R implementation from scratch (without MCMC libraries), detailing acceptance ratios and proposal step-size calibration.
- [Portfolio Risk Model](02-risk-application.qmd): Sampling the 10-factor posterior under fixed covariance to evaluate Value at Risk (VaR) and Expected Shortfall (ES), benchmarked against the exact closed-form posterior.
- [Audit & Diagnostics](03-reproducibility.qmd): Four-chain convergence checks with split-$\hat{R}$, autocorrelation decay, and Effective Sample Size (ESS) audits cross-checked against the `posterior` package.



